---
title: |
  | Green Taxi: 
  | Competitor Market Analysis
subtitle: "Assessed By `r params$assessor`"
author: |
  | Report Generated By `r params$generator`
  | Written By James Wright
date: |
  | Report Generated On
  | `r format(Sys.Date(), format="%A %d %B %Y")`
  | 
  | Date First Published 
  | `r format(as.Date("2023-06-07"), format="%A %d %B %Y")`
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: kable
params:
  generator: 
    label: "Person Generating The Report:"
    value: "User"
  assessor:
    label: "Peson Assessing The Report:"
    value: "Barbara Mikulášová"
  dir:
    label: "Path to Root Folder of Project:"
    value: "/Users/jameswright/Desktop/Amadeus/Graduate Programme/katalyze-taxi-case-study/"
---

\pagebreak

```{r include=FALSE}
knitr::opts_chunk$set( 
  root.dir = getwd(),
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE, 
  include = FALSE)

path <- params$dir
programs_path <- file.path(path, 'R', 'Programs')
source(file.path(path, 'R', 'autoexec.R'), local = knitr::knit_global())
source(file.path(programs_path, 'data_import.R'), local = knitr::knit_global())
```

# Data Collection and Quality Report

```{r}
source(file.path(programs_path, 'data_processing.R'), local = knitr::knit_global())
```

## Data Collection

### Metadata Quality Issues

Summary of problems and corrections with the metadata on import:

* All variable names should be upper-case by convention for formatting clarity;
* Inconsistent variable name word-separation convention across table variables - variable PAYMENTTYPE in taxi_trips renamed to PAYMENT_TYPE ;
* Inconsistent ID variable name convention and naming - variable LOCATION_CODEID in taxi_time_location renamed to LOCATIONID for a consistent ID format across tables and common name across tables;
* Missing variables - variable LOCATION_DETAILS column in taxi_time_location was missing but is in appendix- added containing NA values for completeness.
* Ill-defined field categories in appendix, 'no charge', 'dispute', 'voided' need definitions.

### Duplication Issues

There were `r nrow(taxi_trips_duplicates)` duplicates in the taxi_trips data, `r nrow(taxi_zone_lookup_duplicates)` duplicates in the taxi_zone_lookup data, and `r nrow(taxi_time_location_duplicates)` duplicates in the taxi_time_location data. The duplicates have been saved as an Excel file in the 'Reports' folder to review which data gets duplicated in the collection process.

### Unexpected Data Value Issues
* Unexpected values; 
  + In rows of taxi_zone_lookup, zone 'NV' inconsistent format with other zones. Treated as typo for NA.
* Unexpected categories;
  + Payment type 7 is not defined in the appendix but exists in the data. 
    - For analysis purposes, these were redefined as NA as all corresponded to Vendor 3 which has mostly unknown data.
  + Rate code 8 is not defined in the appendix but exists in the data. Leaving as it is, as all correspond to vendor 3 who might have their own rate code like London cabs.
  + There is a category unknown for payment type - just leave these as NA for convention.

### Missing Data Issues

* Inconsistency with naming convention for missing values - e.g. Unknown, N/A, NA, NV. 
  + All not available data should follow default naming convention NA.
* The entire Ehail column is missing. Data must be collected or stop storing the variable.
    - For analysis purposes, the column was dropped as it was not needed.
* The forward flag, 

\newpage

#### Count of Columns with Missing Values
```{r include=TRUE}
# Display Only Columns with Missing Values and Their Count
kable(taxi_trips_unduplicated_number_missing %>% 
        select_if(~ all(.>0)) %>% 
        names(c("Passenger Count", "E-Hail Fee", "Payment Type", "Congestion Charge")))
```

### Extreme Data Issues

#### Unduplicated Data Summary Report


\pagebreak
# Data Analysis Report